A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-25 paper.tex
eARG conversion tweaks

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs restrospective ARGs}
ARGs are usually thought of as a restrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic inheritance must have occured between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
we have no information about it, given these sampled genomes.
Event ARGs, given their roots in the coalescent with recombination,
are inherently retrospective. They define the evolutionary
events that have occured in the genetic ancestry of a sample.
% This is weak, rephrase.
An event ARG must have some set of samples, or there
will be nothing to define events in terms of.

@@

Prospective ARGs occur in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree'').
Describing this instead as a ``prospective ARG'', and emphasising
the use of the same data structures and direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm which converts a prospective
gARG into a retrospective one is a powerful tool,
with several other applications
as we discuss in the following sections.

@@

Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
(using the ``simplify'' algorithm), and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph. The ``grand MRCA'' $q$ is ommitted from the
fully resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between $g$ and $j$ is omitted because the recombination
event at position $5$ fell in non-ancestral material.
(This example of an eARG
from \citet{wiuf1999recombination} is something of a straw-man,
in that it is highly unlikely that an inference or simulation
program would produce an ARG that contains nodes or edges
that are not ancestral to the sample.)
More generally, the gARG encoding allows for ``local'' inspection
(with respect to a set of samples)
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the effects
of multiple events can be reasoned about at once. Many algorithms
that we wish to perform on an ARG will require resolving
the genetic material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this once and to store the result, whereas
the eARG encoding requires us to repeat the process.

>>>>

% \section*{Inheritance and ancestral material}\label{Ancestry_resolution}
\section*{Prospective vs restrospective ARGs}
ARGs are usually thought of as a restrospective structure, in which we
reason about the genomes that are genetic ancestors of some set of
sample genomes. Figure~\ref{fig-arg-in-pedigree}
is retrospective because it only includes information
about genetic inheritance relative to the sample individuals
$I_1$ and $I_2$ (genomes \textsf{a}--\textsf{d}).
Genetic inheritance must have occured between (e.g.)
$I_8$ and her daughter $I_6$, but this is not recorded because
% this is probably not true, given we know the complete pedigree?
we have no information about it, given these samples.
Event ARGs, given their roots in the coalescent with recombination,
are inherently retrospective. They define the evolutionary
events that have occured in the genetic ancestry of a sample.
% This is weak, rephrase.
An event ARG must have some set of samples, or there
will be nothing to define events in terms of.

@@

Prospective ARGs occur in individual-based forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree},
where the genetic inheritance between genomes is recorded
exhaustively and periodically
``simplified'' to retain only
the ancestry relevant to the current population.
\citet{kelleher2018efficient} refer to the structure
recording all such forwards-in-time genetic relationships
as an ``embellished pedigree`` (and suggest the term ``nedigree''
as a shorthard).
Describing this instead as a ``prospective ARG'', emphasising
the use of the same data structures and the direct connection with
the usual retrospective interpretation of an ARG, may be
more useful.
The ``simplify'' algorithm which converts a prospective
gARG into a retrospective one is a powerful tool,
with several other applications
as we discuss in the following sections.

@@

Fig~\ref{fig-ancestry-resolution}C shows the graph that we obtain
by resolving the ancestral material transmitted along each edge
(using the ``simplify'' algorithm), and highlights some important
differences between the eARG and gARG encodings.
Firstly, we can see that some nodes and edges are removed entirely
from the graph. The ``grand MRCA'' $q$ is ommitted from the
fully resolved gARG because all segments of the genome have
fully coalesced before it is reached. Likewise, the edge
between $g$ and $j$ is omitted because the recombination
event at position $5$ fell in non-ancestral material.
More generally, we can see that the fully resolved
gARG encoding of Fig~\ref{fig-ancestry-resolution}C
allows for ``local'' inspection
of an ARG in a way that is not possible by storing
common ancestor and recombination events in an eARG. Because
the ancestral material is stored with each edge, the effects
of multiple events can be reasoned about at once. Many algorithms
that we wish to perform on an ARG will require resolving
the genetic material with respect to a sample.
% this is weak, but would be good to emphasise this point.
% you just end up running simplify lots of times, in your
% downstream algs
The gARG encoding
allows us to do this once and to store the result, whereas
the eARG encoding requires us to repeat the process on
each traversal.

It is important to note here that the \citet{wiuf1999recombination} eARG
example used here is something of a straw man, in that it is highly unlikely
that an inference or simulation program would produce a retrospective
ARG that contains
nodes or edges that are not ancestral to the sample. Nonetheless, it
is an instructive example from the literature which highlights several
important properties of ARGs, and the general point about
the need to resolve ancestral material ``on the fly'' for eARG traversals
holds.
